### 2022-06-09 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_networkary_classifiers@e4608d1df93d7206dc91c3c754a12d9a884e5ac0 🚀

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily Guidelines for the validation of machine
learning predictions of species interactions}
\vskip 5em

@@

\begin{enumerate}
    \item The prediction of species interactions is gaining momentum as
a way to circumvent limitations in data volume. Yet, ecological networks
are challenging to predict because they are typically small and sparse.
Dealing with extreme class imbalance is a challenge for most binary
classifiers, and there are currently no guidelines as to how predictive
models can be trained for this specific problem.%
    \item Using simple mathematical arguments and numerical experiments
in which a variety of classifiers (for supervised learning) are trained
on simulated networks, we develop a series of guidelines related to the
choice of measures to use for model selection, and the degree of
unbiasing to apply to the training dataset.%
    \item Neither classifier accuracy nor the ROC-AUC are informative
measures for the performance of interaction prediction. PR-AUC is a
fairer assessment of performance. In some cases, even standard measures
can lead to selecting a more biased classifier because the effect of
connectance is strong. The amount of correction to apply to the training
dataset depends on network connectance, on the measure to be optimized,
and only weakly on the classifier.%
    \item These results reveal that training machines to predict
networks is a challenging task, and that in virtually all cases, the
composition of the training set needs to be experimented on before
performing the actual training. We discuss these consequences in the
context of the low volume of data.%
\end{enumerate}

@@

\begin{figure}
\hypertarget{fig:optimvalue}{%
\centering
\includegraphics{figures/optimal_value.png}
\caption{When trained on their optimally biased training set, most
models were able to maximize their performance; this is not true for
decision tree, k-NN, and to a lower extent RF. The ensemble had a
consistently high performance despite incorporating poor
models.}\label{fig:optimvalue}
}
\end{figure}

@@

The trained models were then thresholded (again by optimising
informedness), and their predictions transformed back into networks for
analysis; specifically, we measured the connectance, nestedness (REF),
modularity (REF), asymmetry, and network dissimilarity (REF). This
process was repeated 250 times, and the results are presented in
tbl.~\ref{tbl:comparison}. The random forest model is an interesting
instance here: it produces the network that looks the most like the
original dataset, despite having a very low PR-AUC, suggesting it hits
high recall at the cost of low precision. Although the ensemble was able
to reach a very high PR-AUC (and a very high ROC-AUC), this did not
necessarily translate into more accurate reconstructions of the
structure of the network. This result bears elaborating. Measures of
model performance capture how much of the interactions and
non-interactions are correctly identified. As long as these predictions
are not perfect, some interactions will be predicted at the ``wrong''
position in the network; these measures cannot describe the structural
effect of these mistakes. On the other hand, measures of network
structure can have the same value with interactions that fall at
drastically different positions; this is in part because a lot of these
measures covary with connectance, and in part because as long as these
values are not 0 or their respective maximum, there is a large number of
network configurations that can have the same value. That ROC-AUC is
consistently larger than PR-AUC may be a case of this measure masking
models that are not, individually, strong predictors (Jeni et al.,
2013).

@@

It is noteworthy that the ensemble model was systematically better than
the component models. We do not expect that ensembles will \emph{always}
be better than single models. Networks with different structures than
the one we simulated here may respond in different ways, especially if
the rules are fuzzier than the simple rule we used here. In a recent
multi-model comparison involving supervised and unsupervised learning,
Becker et al. (2022) found that the ensemble was \emph{not} the best
model. There is no general conclusion to draw from this besides
reinforcing the need to be pragmatic about which models should be
included in the ensemble, or whether to use an ensemble at all. In a
sense, the surprising performance of the ensemble model should form the
basis of the last recommendation: optimal training set balance and its
interaction with connectance and the specific binary classifier used is,
in a sense, an hyperparameter that should be assessed. The distribution
of results in fig.~\ref{fig:optimbias} and fig.~\ref{fig:optimvalue}
show that there are variations around the trend. TODO fix this sentence

We establish that due to the low prevalence of interactions, even poor
classifiers applied to food web data will reach a high accuracy; this is
because the measure is dominated by the accidentally correct predictions
of negatives. On simulated confusion matrices with ranges of imbalance
that are credible for ecological networks, MCC had the most desirable
behavior, and informedness is a linear measure of classifier skill. By
performing simulations with four models and an ensemble, we show that
informedness and ROC-AUC are consistently high on network data, and that
MCC and PR-AUC are more accurate measures of the effective performance
of the classifier. Finally, by measuring the structure of predicted
networks, we highlight an interesting paradox: the models with the best
performance measures are not necessarilly the models with the closest
reconstructed network structure. We discuss these results in the context
of establishing guidelines for the prediction of ecological
interactions.

The results presented here highlight an interesting paradox: although
the Random Forest was ultimately able to get a correct estimate of
network structure tbl.~\ref{tbl:comparison}, it ultimately remains a
poor classifier, as evidenced by its low PR-AUC. This suggests that the
goal of predicting \emph{interactions} and predicting \emph{networks}
may not be solvable in the same way -- of course a perfect classifier of
interactions would make a perfect network prediction; but even the best
scoring predictor of interactions (the ensemble model) had not
necessarily the best prediction of network structure. The tasks of
predicting networks structure and of predicting interactions within
networks are essentially two different ones. For some applications
(\emph{.e.g.} comparison of network structure across gradients), one may
care more about a robust estimate of the structure, at the cost at
putting some interactions at the wrong place. For other applications
(\emph{e.g.} identifying pairs of interacting species), one may
conversely care more about getting as many pairs right, even though the
mistakes accumulate in the form of a slightly worse estimate of network
structure. How these two approaches can be reconciled is undoubtedly a
task for further research. Despite this apparent tension at the heart of
the predictive exercise, we can use the results presented here to
suggest a number of guidelines.

First, because we should have more trust in reported interactions than
in reported absences of interactions, we can draw on previous literature
to recommend informedness as a measure to decide on a threshold (Chicco
et al., 2021); this being said, because informedness is insensitive to
bias, the model performance is better evaluated through the use of MCC
fig.~\ref{fig:biasmccinf}. Because \(F_1\) is monotonously sensitive to
classifier bias fig.~\ref{fig:bias} and network connectance
fig.~\ref{fig:connectance}, MCC should be prefered as a measure of model
evaluation and comparison.

Second, because the PR-AUC responds more to network connectance
fig.~\ref{fig:optimvalue} and training set imbalance
fig.~\ref{fig:biasrocpr}, it should be used as a measure of model
performance over the ROC-AUC. This is not to say that ROC-AUC should be
discarded (in fact, a low ROC-AUC is a sign of an issue with the model),
but that its interpretation should be guided by the PR-AUC value.
Specifically, a high ROC-AUC is not informative, as it can be associated
to a low PR-AUC (see \emph{e.g.} Random Forest in
tbl.~\ref{tbl:comparison}) This again echoes recommendations from other
fields (Jeni et al., 2013; Saito \& Rehmsmeier, 2015).

Third, for the same reasons that should give more weight to PR-AUC than
ROC-AUC, accuracy alone should not be used as a measure of model
performance, but rather as an expectation of how well the model should
behave given the balance in the set on which predictions are made; this
is because, as derived earlier, the expected accuracy for a no-skill
no-bias classifier is \(\rho^2 + (1-\rho)^2\) (where \(\rho\) is the
class balance), which will very easily by large. This pitfall is notably
illustrated in a recent model (Caron et al., 2022) wherein the authors
claim that using a training set of \(n = 10^4\) with only 100 positive
interactions (representing 0.1\% of the total interactions) reached a
good accuracy. Reporting a good accuracy isn't informative is this
accuracy isn't (i) compared to the baseline expected value under the
class balance, and (ii) interpreted in the context of a measure that
isn't sensitive to true negatives (like MCC).

Finally, network connectance (\emph{i.e.} the empirical class imbalance)
should inform the composition of the training and testing set. In the
approach outlined here, we treat the class imbalance as an
hyper-parameter, but \emph{test} the model on a set that has the same
class imbalance as the actual dataset. This is an important change to
the overall methodology so far, as it ensure that the prediction
environment matches the testing environment, and so the values on the
testing set can be directly compared to the values for the actual
prediction. A striking result fig.~\ref{fig:optimbias} is that
Informedness was almost always maximal at 50/50 balance (regardless of
connectance), whereas MCC required \emph{more} positives to be maximized
when connectance \emph{increases}. This has an important consequence in
ecological networks, for which the pool of positive cases (interactions)
to draw from is typically small: the most parsimonious measure
(\emph{i.e.} the one requiring to discard the least amount of
information to train the model) will give the best validation potential,
and is probably the informedness (maximizing informedness is the
generally accepted default for imbalanced classification; Schisterman et
al., 2005). This last result further strengthens the assumption that the
amount of bias \emph{is} an hyper-parameter that must be fine-tuned, as
using the wrong bias can lead to models with excess biass or lower
skill.

@@

\textbf{Acknowledgements:} We acknowledge that this study was conducted
on land within the traditional unceded territory of the Saint Lawrence
Iroquoian, Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak
nations. We thank Colin J. Carlson, Michael D. Catchen, Giulio Valentino
Dalla Riva, and Tanya Strydom for inputs on earlier versions of this
manuscript. This research was enabled in part by support provided by
Calcul Québec (www.calculquebec.ca) and Compute Canada
(www.computecanada.ca) through the Narval general purpose cluster. TP is
supported by a NSERC Discovery Grant and Discovery Acceleration
Supplement, by funding to the Viral Emergence Research Initiative
(VERENA) consortium including NSF BII 2021909, and by a grant from the
Institut de Valorisation des Données (IVADO).

>>>>

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily Guidelines for the prediction of species
interactions through binary classification}
\vskip 5em

@@

\begin{enumerate}
    \item The prediction of species interactions is gaining momentum as
a way to circumvent limitations in data volume. Yet, ecological networks
are challenging to predict because they are typically small and sparse.
Dealing with extreme class imbalance is a challenge for most binary
classifiers, and there are currently no guidelines as to how predictive
models can be trained for this specific problem.%
    \item Using simple mathematical arguments and numerical experiments
in which a variety of classifiers (for supervised learning) are trained
on simulated networks, we develop a series of guidelines related to the
choice of measures to use for model selection, and the degree of
unbiasing to apply to the training dataset.%
    \item Neither classifier accuracy nor the ROC-AUC are informative
measures for the performance of interaction prediction. PR-AUC is a
fairer assessment of performance. In some cases, even standard measures
can lead to selecting a more biased classifier because the effect of
connectance is strong. The amount of correction to apply to the training
dataset depends on network connectance, on the measure to be optimized,
and only weakly on the classifier.%
    \item These results reveal that training machines to predict
networks is a challenging task, and that in virtually all cases, the
composition of the training set needs to be fine-tuned before performing
the actual training. We discuss these consequences in the context of the
low volume of data.%
\end{enumerate}

@@

\begin{figure}
\hypertarget{fig:optimvalue}{%
\centering
\includegraphics{figures/optimal_value.png}
\caption{When trained on their optimally biased training set, most
models were able to maximize their performance; this is not true when
measuring PR-AUC for decision tree, k-NN, and to a lower extent RF. The
ensemble had a consistently high performance despite incorporating
low-performing models.}\label{fig:optimvalue}
}
\end{figure}

@@

The trained models were then thresholded (again by optimising
informedness), and their predictions transformed back into networks for
analysis; specifically, we measured the connectance, nestedness
(\(\eta\); \textbf{Bastolla2009ArcMut?}), modularity (\(Q\);
\textbf{Barber2007ModCom?}), asymmetry (\(A\);
\textbf{Delmas2018AnaEco?}), and Jaccard network dissimilarity (Canard
et al., 2014). This process was repeated 250 times, and the results are
presented in tbl.~\ref{tbl:comparison}. The k-NN model is an interesting
instance here: it produces the network that looks the most like the
original dataset, despite having the lowest PR-AUC, suggesting it hits
high recall at the cost of low precision. The ensemble was able to reach
a very high PR-AUC (and a very high ROC-AUC), which translated into more
accurate reconstructions of the structure of the network (with the
exception of modulairty, which is underestimated by \(0.03\)). This
result bears elaborating. Measures of model performance capture how much
of the interactions and non-interactions are correctly identified. As
long as these predictions are not perfect, some interactions will be
predicted at the ``wrong'' position in the network; these measures
cannot describe the structural effect of these mistakes. On the other
hand, measures of network structure can have the same value with
interactions that fall at drastically different positions; this is in
part because a lot of these measures covary with connectance, and in
part because as long as these values are not 0 or their respective
maximum, there is a large number of network configurations that can have
the same value. That ROC-AUC is consistently larger than PR-AUC may be a
case of this measure masking models that are not, individually, strong
predictors (Jeni et al., 2013). In this specif example, the combination
of individually ``adequate'' models resulted in an extremely strong
ensemble, suggesting that the correct prediction of interactions (as
measured by MCC, Inf., ROC-AUC, and PR-AUC) \emph{and} network
properties is indeed a feasible task under appropriately
hyper-parameterized models.

@@

We establish that due to the low prevalence of interactions, even poor
classifiers applied to food web data will reach a high accuracy; this is
because the measure is dominated by the accidentally correct predictions
of negatives. On simulated confusion matrices with ranges of imbalance
that are credible for ecological networks, MCC had the most desirable
behavior, and informedness is a linear measure of classifier skill. By
performing simulations with four models and an ensemble, we show that
informedness and ROC-AUC are consistently high on network data, whereas
MCC and PR-AUC are more accurate measures of the effective performance
of the classifier. Finally, by measuring the structure of predicted
networks, we highlight an interesting paradox: the models with the best
performance measures are not necessarilly the models with the closest
reconstructed network structure. We discuss these results in the context
of establishing guidelines for the prediction of ecological
interactions.

It is noteworthy that the ensemble model was systematically better than
the component models. We do not expect that ensembles will \emph{always}
be better than single models. Networks with different structures than
the one we simulated here may respond in different ways, especially if
the rules are fuzzier than the simple rule we used here. In a recent
multi-model comparison involving supervised and unsupervised learning,
Becker et al. (2022) found that the ensemble was \emph{not} the best
model, and was specifically under-performing compared to models using
biological traits. This may be because the dataset of Becker et al.
(2022) was known to be under-sampled, and so the network itself
contained less information than the network and species traits. There is
no general conclusion to draw from either these results or ours, besides
reinforcing the need to be pragmatic about which models should be
included in the ensemble, and whether to use an ensemble at all. In a
sense, the surprising performance of the ensemble model should form the
basis of the first broad recommendation: optimal training set balance
and its interaction with connectance and the specific binary classifier
used is, in a sense, an hyperparameter that should be assessed. The
distribution of results in fig.~\ref{fig:optimbias} and
fig.~\ref{fig:optimvalue} show that there are variations around the
trend, and multiple models should probably be trained on their
``optimal'' training/testing set, as opposed to the same ones.

The results presented here highlight an interesting paradox: although
the k-NN model was ultimately able to get a correct estimate of network
structure tbl.~\ref{tbl:comparison}, it ultimately remains a poor
classifier, as evidenced by its low PR-AUC. This suggests that the goal
of predicting \emph{interactions} and predicting \emph{networks} may not
always be solvable in the same way -- of course a perfect classifier of
interactions would make a perfect network prediction; indeed, the best
scoring predictor of interactions (the ensemble model) had the best
prediction of network structure. The tasks of predicting networks
structure and of predicting interactions within networks are essentially
two different ones. For some applications (\emph{.e.g.} comparison of
network structure across gradients), one may care more about a robust
estimate of the structure, at the cost at putting some interactions at
the wrong place. For other applications (\emph{e.g.} identifying pairs
of interacting species), one may conversely care more about getting as
many pairs right, even though the mistakes accumulate in the form of a
slightly worse estimate of network structure. How these two approaches
can be reconciled is something to evaluate on a case-by-case basis,
especially since there is no guarantee that an esemble model will always
be the most precise one. Despite this apparent tension at the heart of
the predictive exercise, we can use the results presented here to
suggest a number of guidelines.

First, because we have more trust in reported interactions than in
reported absences of interactions (which are overwhelmingly
\emph{pseudo}-absences), we can draw on previous literature to recommend
informedness as a measure to decide on a threshold for binary
classification (Chicco et al., 2021); this being said, because
informedness is insensitive to bias (although it is a linear measure of
skill), the overall model performance is better evaluated through the
use of MCC fig.~\ref{fig:biasmccinf}. Because \(F_1\) is monotonously
sensitive to classifier bias fig.~\ref{fig:bias} and network connectance
fig.~\ref{fig:connectance}, MCC should be prefered as a measure of model
evaluation and comparison. When dealing with multiple models, we
therefore suggest to find the optimal threshold using informedness, and
to pick the best model using MCC (assuming one does not want to use an
ensemble model).

Second, accuracy alone should not be the main measure of model
performance, but rather an expectation of how well the model should
behave given the class balance in the set on which predictions are made;
this is because, as derived earlier, the expected accuracy for a
no-skill no-bias classifier is \(\rho^2 + (1-\rho)^2\) (where \(\rho\)
is the class balance), which will most often be large. This pitfall is
notably illustrated in a recent food-web model (Caron et al., 2022)
wherein the authors, using a training set of \(n = 10^4\) with only 100
positive interactions (representing 0.1\% of the total interactions),
reached a good accuracy. Reporting a good accuracy is not informative,
especially when accuracy isn't (i) compared to the baseline expected
value under the given class balance, and (ii) interpreted in the context
of a measure that is not sensitive to the chance prediction of many
negatives (like MCC).

Third, because the PR-AUC responds more to network connectance
fig.~\ref{fig:optimvalue} and training set imbalance
fig.~\ref{fig:biasrocpr} than ROC-AUC, it should be used as a measure of
model performance over the ROC-AUC. This is not to say that ROC-AUC
should be discarded (in fact, a low ROC-AUC is undoubtedly a sign of an
issue with the model), but that its interpretation should be guided by
the PR-AUC value. Specifically, a high ROC-AUC is not informative, as it
can be associated to a low PR-AUC (see \emph{e.g.} Random Forest in
tbl.~\ref{tbl:comparison}) This again echoes recommendations from other
fields (Jeni et al., 2013; Saito \& Rehmsmeier, 2015). We therefore
expect to see high ROC-AUC values, and then to pick the model that
maximizes the PR-AUC value. Taken together with the previous two
guidelines, we strongly encourage to (i) ensure that accuracy and
ROC-AUC are high (in the case of accuracy, higher than expected under
no-skill no-bias situation), and (ii) to discuss the performance of the
model in terms of the most discriminant measures, \emph{i.e.} PR-AUC and
MCC.

Finally, network connectance (\emph{i.e.} the empirical class imbalance)
should inform the composition of the training and testing set, because
it is an ecologically relevant value. In the approach outlined here, we
treat the class imbalance of the training set as an hyper-parameter, but
\emph{test} the model on a set that has the same class imbalance as the
actual dataset. This is an important distinction, as it ensure that the
prediction environment matches the testing environment (as we cannot
manipulate the connectance of the empirical dataset on which the
predictions will be made), and so the values measured on the testing set
(or validation set if the data volume allows one to exists) can be
directly compared to the values for the actual prediction. A striking
result fig.~\ref{fig:optimbias} is that Informedness was almost always
maximal at 50/50 balance (regardless of connectance), whereas MCC
required \emph{more} positives to be maximized when connectance
\emph{increases}, matching the idea that it is a more stringent measure
of performance. This has an important consequence in ecological
networks, for which the pool of positive cases (interactions) to draw
from is typically small: the most parsimonious measure (\emph{i.e.} the
one requiring to discard the least amount of interactions to train the
model) will give the best validation potential, and in this light is
very likely informedness (maximizing informedness is, in fact, the
generally accepted default for imbalanced classification regardless of
the problem domain; Schisterman et al., 2005). This last result further
strengthens the assumption that the amount of bias \emph{is} an
hyper-parameter that must be fine-tuned, as using the wrong bias can
lead to models with lower performance; for this reason, it makes sense
to not train all models on the same training/testing set, but rather to
optimize the set composition for each of them.

@@

\textbf{Acknowledgements:} We acknowledge that this study was conducted
on land within the traditional unceded territory of the Saint Lawrence
Iroquoian, Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak
nations. We thank Colin J. Carlson, Michael D. Catchen, Giulio Valentino
Dalla Riva, and Tanya Strydom for inputs on earlier versions of this
manuscript. This research was enabled in part by support provided by
Calcul Québec (www.calculquebec.ca) through the Narval general purpose
cluster. TP is supported by the Fondation Courtois, a NSERC Discovery
Grant and Discovery Acceleration Supplement, by funding to the Viral
Emergence Research Initiative (VERENA) consortium including NSF BII
2021909, and by a grant from the Institut de Valorisation des Données
(IVADO).
